---
title: "MELODEM Workshop Paper (title TBD)"
bibliography: refs.bib
csl: jama.csl
always_allow_html: true
output: 
  officedown::rdocx_document:
    reference_docx: style_manuscript_times_new_roman.docx
    keep_md: true
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE,
                      dpi = 600, 
                      fig.width = 11, 
                      fig.height = 6)

# unified specification for rounding in the paper.
# if |value| is 0 to 1, round to 2 decimals
# if |value| is 1 to 10, round to 1 decimal
# if |value| is > 10, round to 0 decimals
rspec <- round_spec() |>
  round_using_magnitude(digits = c(2, 1, 0),
                        breaks = c(1, 10, Inf))

# save it to options:
names(rspec) <- paste('table.glue', names(rspec), sep = '.')
options(rspec)

# make a short name for table.glue::table_value - this makes
# the inline reporting of results less complicated to read.
tv <- table_value
tg <- table_glue


# default is double space 
# single spacing for title page, abstract, captions
single_par <- fp_par(line_spacing=1)

# default page margins
margins_normal <- page_mar(bottom = 2.54 * 0.3937, 
                           top    = 2.54 * 0.3937, 
                           right  = 2.54 * 0.3937, 
                           left   = 2.54 * 0.3937,
                           gutter = 0)

plain_text <- function(x){
  paste(x[[1]]$txt, collapse = '')
}


```

Your name here, your degree,^1^ 

^1^Your department, University, City, State. `r single_par`

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

**Correspondence to:**\
Byron C. Jaeger, PhD\
Department of Biostatistics and Data Science\
Division of Public Health Sciences\
Wake Forest School of Medicine\
Medical Center Boulevard\
Winston-Salem, NC 27154\
336-716-6956\
[bjaeger\@wakehealth.edu](mailto:bjaeger@wakehealth.edu){.email} `r single_par`

**Word Count:** Abstract: ; Manuscript: 

\newpage

```{r load-targets}

withr::with_dir(
  here::here(),
  code = {
    targets::tar_load(
      names = c("data_sim_1",
                "data_sim_2",
                "data_pooled",
                "bm_risk")
    )
  }
)

```


# ABSTRACT

\newpage

# INTRODUCTION


Here's how to cite one thing.[@r_language]

Here's how to cite two or more things.[@ambrosius_design_2014; @sprint_research_group_randomized_2015]

# METHODS

Chronic kidney disease was defined by `r labels$definitions$ckd`

# RESULTS

# DISCUSSION

\newpage

# REFERENCES

<div id="refs"></div>

\newpage

# ACKNOWLEDGMENTS 

# FINANCIAL DISCLOSURE 

```{r tbl_footers}



```



<!-- new page not needed at the end of a block section -->

```{r}

block_section(
  prop_section(
    page_size = page_size(orient = "portrait"),
    type = "continuous",
    page_margins = margins_normal
  )
)

```

To make a landscape section, use the block_section code here. Note that the section will be applied to all content prior to the block reaching up to the most recent block.

```{r echo = TRUE}

block_section(
  prop_section(
    page_size = page_size(orient = "landscape"),
    page_margins = page_mar(bottom = 1/8, top = 1/8, 
                            right = 1/4, left = 1/4),
    type = "continuous"
  )
)
```


To switch back to regular portrait section, add another portrait block. Notice how the block below acts on all content above it up to the prior block.


```{r echo = TRUE}
block_section(
  prop_section(
    page_size = page_size(orient = "portrait"),
    type = "continuous",
    page_margins = margins_normal
  )
)
```

