---
title: "MELODEM Workshop Paper Results"
bibliography: refs.bib
csl: jama.csl
always_allow_html: true
output: 
  officedown::rdocx_document:
    reference_docx: style_arial.docx
    keep_md: true
    toc: true
    toc_depth: 2
---

```{r setup-knitr, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,    # don't show code blocks
  message = FALSE, # don't print messages to the doc
  warning = FALSE, # don't print warnings to the doc
  dpi = 300,       # higher dpi => higher render time 
  fig.width = 6,   # default figure width in inches
  fig.height = 4   # default figure height in inches
)

library(officedown)
library(officer)

```

```{r setup-figures}

set_fig_landscape <- function() {
  knitr::opts_chunk$set(fig.width = 10.5, fig.height = 6)
}

set_fig_portrait <- function() {
  knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
}

theme_set(
  new = theme_bw() + 
    theme(panel.grid = element_blank(),
          text = element_text(size = 12))
)


```

```{r setup-margins}

# Set a default footer because when we use block_section, it 
# starts a new style template for headers and footers. Thus,
# if we want page numbers to appear on every page, we need to
# put that into the default footer and then put that default
# footer into every block section. MS word...¯\_(ツ)_/¯
# note to future self - go here if you want to remember this: 
# https://github.com/davidgohel/officedown/discussions/103

footer_default <- block_list(
  fpar(
    run_word_field(field = "PAGE"),
    fp_p = fp_par(text.align = "left") 
  )
)

# Code in this chunk sets the margins for your document. 
# The helper functions page_long_above() and page_wide_above()
# also provide shortcuts for switching between wide (landscape)
# and normal (long) MS word layouts. 

# helper function for defining margins
make_margin <- function(x=NULL){
  
  if(is.null(x)){ return(page_mar()) } 
  
  if(is.character(x)){
    
    out <- switch(
      tolower(x),
      "normal" = page_mar(1, 1, 1, 1),
      "narrow" = page_mar(1/2, 1/2, 1/2, 1/2),
      "moderate" = page_mar(1, 1, 3/4, 3/4)
    )
    
  }
  
}

# set content above to be portrait
page_long_above <- function(margins = make_margin('narrow')){
  
  # b/c block_sections apply to content ABOVE, we set
  # the content BELOW to be landscape when we use page_long_above  
  set_fig_landscape()
  
  block_section(
    prop_section(
      page_size = page_size(orient = "portrait"),
      type = "continuous",
      page_margins = margins,
      footer_default = footer_default
    )
  )
  
}

# set content above to be landscape
page_wide_above <- function(margins = make_margin('narrow')){
  
  # b/c block_sections apply to content ABOVE, we set
  # the content BELOW to be portrait when we use page_wide_above
  set_fig_portrait()
  
  block_section(
    prop_section(
      page_size = page_size(orient = "landscape"),
      type = "continuous",
      page_margins = margins,
      footer_default = footer_default
    )
  )
  
}

```

```{r setup-targets}

withr::with_dir(
  new = here::here(),
  code = {
    targets::tar_load(names = c(tbl_characteristics,
                                tbl_het,
                                tbl_coxph,
                                fig_incidence))
  }
)

```


```{r setup-tbl}

# Footnotes should be wrapped in `as_paragraph`:

ckd_defn <- as_paragraph("Chronic Kidney Disease was defined by an estimated glomerular filtration rate <60 ml/min/1.73 m", as_sup(2L), " based on the 2021 CKD-EPI creatinine equation.")

# abbreviation management

abbrvs <- list(
  "SD"   = 'standard deviation',
  "ACR" =	"Albumin-to-creatinine ratio",
  "ASCVD" =	"Atherosclerotic cardiovascular disease",
  "CI" =	"Confidence interval",
  "COPD" =	"Chronic obstructive pulmonary disease",
  "eGFR" =	"Estimated glomerular filtration rate",
  "GLP-1RA" =	"Glucagon-like peptide-1 receptor agonist ",
  "HbA1c" =	"Glycated hemoglobin",
  "HDL" =	"High-density lipoprotein",
  "hsCRP" =	"High-sensitivity C-reactive protein",
  "NHANES" =	"National Health and Nutrition Examination Survey",
  "PHQ" =	"Patient Health Questionnaire"
)



abbrvs_paste <- function(strings){
  
  front <- 'Abbreviations'
  
  matter <- glue::glue_collapse(strings, sep = '; ', last = '; and ')
  
  paste(front, matter, sep = ': ')
  
}

abbrvs_write <- function(abbr){
  
  sorted <- abbr[sort(names(abbr))]
  
  strings <- map2_chr(names(sorted), sorted, paste, sep = ' = ')
  
  as_paragraph(abbrvs_paste(strings))
  
}


```

\newpage

## Table 1: Participant characteristics

```{r}
tbl_characteristics
```

\newpage

## Table 2: Hazard ratios for dementia overall and in subgroups based on Apolipoprotein E. 

```{r}
tbl_coxph
```


`r page_long_above()`

## Table 3: Heterogeneity in the effect of APOE-4 on dementia risk using time since birth

```{r}
tbl_het[[1]] %>% 
  flextable_autofit(width_max = 11)
```

\newpage

## Table 3: Heterogeneity in the effect of APOE-4 on dementia risk using time since baseline

```{r}
tbl_het[[2]] %>% 
  flextable_autofit(width_max = 11)
```

## Cumulative incidence of dementia stratified by study

```{r}
fig_incidence
```

`r page_wide_above()`




